---
title: "Getting Started with AccumulatR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with AccumulatR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This article is a short orientation to the main objects and column names used
throughout `AccumulatR`. It is meant to answer the first questions most users
have before they start fitting specific models.

```{r}
library(AccumulatR)
```

## The basic workflow

Most analyses in `AccumulatR` follow the same four steps:

1. Define a model with accumulators, pools, and observed responses.
2. Finalize that model with `finalize_model()`.
3. Work with behavioral data, either simulated with `simulate()` or collected
   in an experiment.
4. Prepare the data, build a model context, and evaluate candidate parameter values.

## Core ideas

### Trial

A `trial` is a single behavioral observation. In the simplest case, each row of
your data corresponds to one trial.

### `R`

`R` is the observed response label on a trial. For a two-choice task, typical
values might be `"left"` and `"right"`, or `"go"` and `"stop"`.

### `rt`

`rt` is the observed response time for `R`.

### Component

A `component` is a latent submodel inside a mixture model. This is useful when
you think behavior comes from qualitatively different trial types, such as:

- a fast-guess process on some trials
- a more controlled decision process on others

If your model has only one component, you usually do not need to think about
this column at all.

### Onset

An `onset` is the time at which an accumulator becomes active. Some
accumulators start at time 0, while others begin later. In `AccumulatR`, an
onset can be:

- a fixed delay, such as `onset = 0.15`
- a chained onset, such as `onset = after("cue")`, meaning the process starts
  only after another accumulator or pool has finished

Although onset and non-decision time (`t0`) have the same role on the response times
We make the distinction that `t0` is an estimated parameter, whereas onset is not.


## A minimal model

The example below defines a basic two-choice race model.

```{r}
spec <- race_spec() |>
  add_accumulator("left", "lognormal") |>
  add_accumulator("right", "lognormal") |>
  add_outcome("left", "left") |>
  add_outcome("right", "right")

model <- finalize_model(spec)
```

## Simulated behavioral data

We can generate a small behavioral dataset from this model.

```{r}
pars <- c(
  left.m = log(0.28), left.s = 0.16, left.q = 0, left.t0 = 0,
  right.m = log(0.35), right.s = 0.18, right.q = 0, right.t0 = 0
)

param_df <- build_param_matrix(spec, pars, n_trials = 6)
sim <- simulate(model, param_df, seed = 123)

head(sim)
```

Here each row is one trial, `R` is the observed response, and `rt` is the
response time.

## From behavioral data to likelihood

To fit a model, the usual pattern is:

```{r}
prepared <- prepare_data(model, sim[c("trial", "R", "rt")])
ctx <- make_context(model)
log_likelihood(ctx, prepared, param_df)
```

`prepare_data()` expands and normalizes the behavioral data for likelihood
evaluation, and `make_context()` compiles the model-side runtime state.
`log_likelihood()` then tells you how well a parameter set explains those
observed data.

## Where to go next

- If you want the basic fitting workflow, read the simple race-model vignette.
- If your experiment records multiple ordered responses, read the multi-outcome
  vignette.
- If your model has staged processing, read the chained-onset vignette.
