---
title: "Getting Started with AccumulatR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with AccumulatR}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

This vignette defines the main workflow used throughout `AccumulatR`.
Read it first if you want the package vocabulary before moving to the
task-specific examples.

```{r}
library(AccumulatR)
```

## The basic workflow

Most workflows in `AccumulatR` follow the same four steps:

1.  Define a model with accumulators, pools, and observed responses.
2.  Finalize that model with `finalize_model()`.
3.  Work with behavioral data, either simulated with `simulate()` or
    collected in an experiment.
4.  Prepare the data, build a model context, and evaluate candidate
    parameter values.

## Core ideas

### Trial

A `trial` is a single behavioral observation. In the simplest case, each
row of your data corresponds to one trial.

### `R`

`R` is the observed response label on a trial. For a two-choice task,
typical values might be `"left"` and `"right"`, or `"red"` and
`"green"`, etc.

### `rt`

`rt` is the observed response time for `R`.

### Component

A `component` serves two purposes. It can be used to account for
different trial types in your data, for example when a subset of your
data requires different accumulator structures. It can also be used to
allow for latent mixtures *within* trial. If your model has only one
trial type, you usually do not need this column. See also the **Working
with Mixtures** vignette.

### Onset

An `onset` is the time at which an accumulator becomes active. Most
accumulators start at time `0`, while in some cases, extra information
appears during the trial. In `AccumulatR`, an onset can be:

-   a fixed delay, such as `onset = 0.15`
-   a chained onset, such as `onset = after("A")`, meaning the process
    starts only after another accumulator (in this case `A`) or pool has
    finished

Both onset and non-decision time (`t0`) shift response times. The
difference in `AccumulatR` is that `t0` is an estimated parameter,
whereas onset is part of the model structure.

## A minimal model

The example below defines a basic two-choice race model with two
lognormal accumulators.

```{r}
model <- race_spec() |>
  add_accumulator("left", "lognormal") |>
  add_accumulator("right", "lognormal") |>
  add_outcome("left", "left") |>
  add_outcome("right", "right") |>
  finalize_model()
```

## Simulated behavioral data

We can generate a small behavioral dataset from this model. Here
`build_param_matrix` makes one row of parameters per trial. This allows
you to trial-wise manipulate the specific parameters, for example using
a design matrix specification.

```{r}
pars <- c(
  left.m = log(0.28), left.s = 0.16,
  right.m = log(0.35), right.s = 0.18
)

param_df <- build_param_matrix(model, pars, n_trials = 6)
sim <- simulate(model, param_df, seed = 123)

head(sim)
```

Each row is one trial, `R` is the observed response, and `rt` is the
response time.

## From behavioral data to likelihood

To fit a model, the usual pattern is:

```{r}
prepared <- prepare_data(model, sim[c("trial", "R", "rt")])
ctx <- make_context(model)
log_likelihood(ctx, prepared, param_df)
```

`prepare_data()` reshapes the behavioral data for likelihood evaluation,
and `make_context()` builds the model-side runtime state.
`log_likelihood()` returns the fit of a parameter set to those observed
data.

## Where to go next

-   If you want the basic fitting workflow, read the simple race-model
    vignette.
-   If you want to learn more about the different structures, read the
    logical rules vignettes.
-   If you want pools of accumulators, read the pool vignette.
-   If you want to include trigger failures, read the trigger vignette.
-   If you want to include mixtures, read the mixtures vignette.
-   If your experiment records multiple ordered responses, read the
    multi-outcome vignette.
-   If your model has staged processing, read the chained-onset
    vignette.
