---
title: "Logical Rules for Responses"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Logical Rules for Responses}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette explains how `AccumulatR` lets you define observed responses with
logical rules. The goal is not just to show the syntax, but to make clear what
the model is assuming when you write rules with `first_of()`, `all_of()`,
`none_of()`, and `inhibit()`.

```{r}
library(AccumulatR)
```

## What a response rule means

In `AccumulatR`, an outcome rule says when an observed response should count as
having happened.

Sometimes a response is a direct readout from a single accumulator. But many
psychological models are more structured than that. A response may depend on:

- one of several routes being able to generate it
- several processes all being complete
- the absence of some competing event
- a stop-like process that can cancel an otherwise available response

The logical helpers are there to describe those ideas directly.

## The four main helpers

It helps to keep one plain-language summary in mind:

- `first_of(x, y)` means the response can be generated by either `x` or `y`
- `all_of(x, y)` means the response requires both `x` and `y`
- `none_of(x)` means the response rule requires `x` not to happen
- `inhibit(x, y)` means `x` would generate the response, but `y` can block it


## `first_of()`: either route can generate the response

Use `first_of()` when an observed response can be produced by one of several
routes.

For example, suppose a participant can detect a target either from a word cue
or from a picture cue. Then the response can be generated by whichever route
finishes first.

```{r}
spec_first <- race_spec() |>
  add_accumulator("word_route", "lognormal") |>
  add_accumulator("picture_route", "lognormal") |>
  add_outcome("target_detected", first_of("word_route", "picture_route"))

finalize_model(spec_first)$outcomes[[1]]$label
```

## `all_of()`: several things must all be true

Use `all_of()` when an observed response depends on several processes all being
complete.

For example, suppose a response should only count once a stimulus has been
encoded and a decision has been verified.

```{r}
spec_all <- race_spec() |>
  add_accumulator("encoding", "lognormal") |>
  add_accumulator("verification", "lognormal") |>
  add_outcome("confirmed_response", all_of("encoding", "verification"))

finalize_model(spec_all)$outcomes[[1]]$label
```

## `none_of()`: the response requires the competing event to be absent

Use `none_of()` when the response rule itself says that some event must not
happen.

For example, suppose a participant makes a response only if a `go` process
finishes and no `stop` event occurs.

```{r}
spec_none <- race_spec() |>
  add_accumulator("go", "lognormal") |>
  add_accumulator("stop", "lognormal") |>
  add_outcome("respond", all_of("go", none_of("stop")))

finalize_model(spec_none)$outcomes[[1]]$label
```

The key assumption here is:

- the response requires `go`
- the response also requires that `stop` not happen

So `stop` is being treated as a condition whose absence is part of the rule.
This is the better description when your theory is literally "respond only on
trials where the stop event did not occur."

## `inhibit()`: the response is driven by one process and blocked by another

Use `inhibit()` when one process would normally generate the response, but
another process can cancel it.

This is the natural way to write stop-signal-like logic.

```{r}
spec_inhibit <- race_spec() |>
  add_accumulator("go", "lognormal") |>
  add_accumulator("stop", "lognormal") |>
  add_outcome("respond", inhibit("go", "stop"))

finalize_model(spec_inhibit)$outcomes[[1]]$label
```

The key assumption here is different:

- `go` is the response-generating process
- `stop` is not just absent or present
- `stop` has the specific role of blocking the response if it gets there first

So `inhibit()` is the better description when your theory is "the go response
would have happened, but a stop process cancelled it."

## `none_of()` versus `inhibit()`

These two rules look similar:

```{r}
rule_none <- all_of("go", none_of("stop"))
rule_inhibit <- inhibit("go", "stop")

rule_none
rule_inhibit
```

But they are not the same model statement.

### `all_of("go", none_of("stop"))`

This says that the response is defined by two conditions:

1. `go` happened
2. `stop` did not happen

So the model treats `stop` as something whose absence is part of the response
definition itself.

### `inhibit("go", "stop")`

This says that:

1. `go` is the event that would generate the response
2. `stop` can prevent that response by beating it

So the model treats `stop` as a blocker of an otherwise available response.

## A timing-based way to think about the difference

The difference becomes clearer if you imagine response times.

Suppose:

- `go` finishes at 300 ms
- `stop` finishes at 250 ms

With `inhibit("go", "stop")`, the response is blocked, because the stop
process won the race to veto the response.

Now suppose instead:

- `go` finishes at 300 ms
- `stop` finishes at 400 ms

With `inhibit("go", "stop")`, the response is not blocked, because `go`
finished first.

That is why `inhibit()` is fundamentally about temporal blocking.

By contrast, `none_of("stop")` is not saying that `stop` raced with `go` and
lost or won. It is saying that the rule requires `stop` not to be part of what
counts for that response.

So:

- `inhibit()` means "stop can block go if it gets there in time"
- `none_of()` means "the response requires stop not to occur"

## A combined example

The helpers can be combined when a response depends on several kinds of logic
at once.

Suppose a response requires:

- an encoding stage to finish
- either a left route or a right route to generate the choice
- no stop event

```{r}
spec_combo <- race_spec() |>
  add_accumulator("encoding", "lognormal") |>
  add_accumulator("left_route", "lognormal") |>
  add_accumulator("right_route", "lognormal") |>
  add_accumulator("stop", "lognormal") |>
  add_outcome(
    "choice",
    all_of(
      "encoding",
      first_of("left_route", "right_route"),
      none_of("stop")
    )
  )

finalize_model(spec_combo)$outcomes[[1]]$label
```

This kind of rule is useful when the observed response is not just "whichever
accumulator wins," but instead reflects a structured decision
rule.

## Choosing the right helper

As a practical guide:

- use `first_of()` when several routes can each produce the same response
- use `all_of()` when several conditions must all be satisfied
- use `none_of()` when the response rule explicitly requires some event to be absent
- use `inhibit()` when one process would produce the response and another process can cancel it

In cognitive-process terms, `none_of()` is about absence as a condition,
whereas `inhibit()` is about cancellation of an otherwise available response.
