---
title: "Logical Rules for Responses"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Logical Rules for Responses}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette shows how to define observed responses with `first_of()`,
`all_of()`, `none_of()`, and `inhibit()`. Use these helpers when a response is
not a direct readout from a single accumulator.

```{r}
library(AccumulatR)
```

## What a response rule does

An outcome rule states when an observed response should count as having
happened. Common cases include:

- one of several routes being able to generate it
- several processes all being complete
- the absence of some competing event
- a stop-like process that can cancel an otherwise available response

The logical helpers encode those cases directly in the model definition.

## The four main helpers

The main helpers are:

- `first_of(x, y)` means the response can be generated by either `x` or `y`
- `all_of(x, y)` means the response requires both `x` and `y`
- `none_of(x)` means the response rule requires `x` not to happen
- `inhibit(x, y)` means `x` would generate the response, but `y` can block it


## `first_of()`: either route can generate the response

Use `first_of()` when one of several routes can generate the same response.

```{r}
model_first <- race_spec() |>
  add_accumulator("word_route", "lognormal") |>
  add_accumulator("picture_route", "lognormal") |>
  add_outcome("target_detected", first_of("word_route", "picture_route")) |>
  finalize_model()

model_first$outcomes[[1]]$label
```

## `all_of()`: several things must all be true

Use `all_of()` when a response depends on several processes all being complete.

```{r}
model_all <- race_spec() |>
  add_accumulator("encoding", "lognormal") |>
  add_accumulator("verification", "lognormal") |>
  add_outcome("confirmed_response", all_of("encoding", "verification")) |>
  finalize_model()

model_all$outcomes[[1]]$label
```

## `none_of()`: the response requires the competing event to be absent

Use `none_of()` when the rule itself requires an event not to happen.

```{r}
model_none <- race_spec() |>
  add_accumulator("go", "lognormal") |>
  add_accumulator("stop", "lognormal") |>
  add_outcome("respond", all_of("go", none_of("stop"))) |>
  finalize_model()

model_none$outcomes[[1]]$label
```

Here the absence of `stop` is part of the response definition.

## `inhibit()`: the response is driven by one process and blocked by another

Use `inhibit()` when one process would generate the response and another can
block it.

```{r}
model_inhibit <- race_spec() |>
  add_accumulator("go", "lognormal") |>
  add_accumulator("stop", "lognormal") |>
  add_outcome("respond", inhibit("go", "stop")) |>
  finalize_model()

model_inhibit$outcomes[[1]]$label
```

Here `go` is the response-generating process and `stop` blocks it.

## `none_of()` versus `inhibit()`

These two rules look similar:

```{r}
rule_none <- all_of("go", none_of("stop"))
rule_inhibit <- inhibit("go", "stop")

rule_none
rule_inhibit
```

But they are not the same model statement.

### `all_of("go", none_of("stop"))`

This defines the response in terms of two conditions: `go` happened and `stop`
did not happen.

### `inhibit("go", "stop")`

This defines `go` as the response-generating event and `stop` as a blocker of
that event.

In short:

- `inhibit()` means "stop can block go if it gets there in time"
- `none_of()` means "the response requires stop not to occur"

## A combined example

The helpers can be combined when a response depends on several kinds of logic
at once. For example, suppose a response requires:

- an encoding stage to finish
- either a left route or a right route to generate the choice
- no stop event

```{r}
model_combo <- race_spec() |>
  add_accumulator("encoding", "lognormal") |>
  add_accumulator("left_route", "lognormal") |>
  add_accumulator("right_route", "lognormal") |>
  add_accumulator("stop", "lognormal") |>
  add_outcome(
    "choice",
    all_of(
      "encoding",
      first_of("left_route", "right_route"),
      none_of("stop")
    )
  ) |>
  finalize_model()

model_combo$outcomes[[1]]$label
```

Use combined rules when the observed response reflects a structured decision
rule rather than a single winning accumulator.

## Choosing the right helper

As a practical guide:

- use `first_of()` when several routes can each produce the same response
- use `all_of()` when several conditions must all be satisfied
- use `none_of()` when the response rule explicitly requires some event to be absent
- use `inhibit()` when one process would produce the response and another process can cancel it

`none_of()` treats absence as part of the response definition. `inhibit()`
treats one process as blocking another.
