---
title: "Working with Mixtures"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with Mixtures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Here we highlight how fixed and sampled mixtures can be instantiated in
`AccumulatR`.

Use fixed components when trial type is known and should be supplied in the
data. Use sampled mixtures when the component is latent and the likelihood
must average over the possible processing modes.

```{r}
library(AccumulatR)
```

## A common architecture

Both sections use the same fast-versus-slow architecture. Response `R1` comes
from a target route and response `R2` comes from a competitor. The components
differ only in whether the target route is fast or slow on a given trial.

## Fixed components: known trial types

In this version, component labels are observed and included in the data.

```{r}
model_fixed <- race_spec() |>
  add_accumulator("target_fast", "lognormal") |>
  add_accumulator("target_slow", "lognormal") |>
  add_accumulator("competitor", "lognormal") |>
  add_pool("TARGET", c("target_fast", "target_slow")) |>
  add_outcome("R1", "TARGET") |>
  add_outcome("R2", "competitor") |>
  add_component("fast", members = c("target_fast", "competitor"), weight = 0.4) |>
  add_component("slow", members = c("target_slow", "competitor"), weight = 0.6) |>
  set_mixture_options(mode = "fixed") |>
  finalize_model()

true_params_fixed <- c(
  target_fast.m = log(0.25),
  target_fast.s = 0.15,
  target_slow.m = log(0.45),
  target_slow.s = 0.20,
  competitor.m = log(0.35),
  competitor.s = 0.18
)
```

We now create a trial table that says which component each trial belongs to.

```{r}
set.seed(123456)

n_trials_fixed <- 1200
trial_types <- data.frame(
  trial = seq_len(n_trials_fixed),
  component = sample(c("fast", "slow"), n_trials_fixed, replace = TRUE, prob = c(0.4, 0.6)),
  stringsAsFactors = FALSE
)

params_df_fixed <- build_param_matrix(
  model_fixed,
  true_params_fixed,
  trial_df = trial_types
)

sim_fixed <- simulate(
  model_fixed,
  params_df_fixed,
  trial_df = trial_types,
  layout = "long"
)

data_fixed <- sim_fixed[, c("trial", "R", "rt", "component")]

table(data_fixed$component, data_fixed$R)
```

Because the component is known, the simulated data keep that label. The same
pattern applies to real data: if trial type is known, keep it in a
`component` column.

```{r}
prepared_fixed <- prepare_data(model_fixed, data_fixed)
ctx_fixed <- make_context(model_fixed)

params_df_fixed_true <- build_param_matrix(
  model_fixed,
  true_params_fixed,
  trial_df = prepared_fixed
)

ll_fixed <- as.numeric(log_likelihood(ctx_fixed, prepared_fixed, params_df_fixed_true))
ll_fixed
```

## Sampled components: latent processing modes

Now we use the same architecture as a latent mixture. The component is not
observed per trial. Instead, the model samples a hidden component for each
trial, with `p_fast` controlling the proportion of fast trials.

```{r}
model_sampled <- race_spec() |>
  add_accumulator("target_fast", "lognormal") |>
  add_accumulator("target_slow", "lognormal") |>
  add_accumulator("competitor", "lognormal") |>
  add_pool("TARGET", c("target_fast", "target_slow")) |>
  add_outcome("R1", "TARGET") |>
  add_outcome("R2", "competitor") |>
  add_component("fast", members = c("target_fast", "competitor"), weight_param = "p_fast") |>
  add_component("slow", members = c("target_slow", "competitor")) |>
  set_mixture_options(mode = "sample", reference = "slow") |>
  finalize_model()

true_params_sampled <- c(
  target_fast.m = log(0.25),
  target_fast.s = 0.15,
  target_slow.m = log(0.45),
  target_slow.s = 0.20,
  competitor.m = log(0.35),
  competitor.s = 0.18,
  p_fast = 0.35
)
```

By default, sampled mixtures do not expose the component label in the output,
because that label is latent.

```{r}
set.seed(123456)

n_trials_sampled <- 1500
params_df_sampled <- build_param_matrix(
  model_sampled,
  true_params_sampled,
  n_trials = n_trials_sampled
)

sim_sampled <- simulate(model_sampled, params_df_sampled)
names(sim_sampled)
```

If you want to inspect the hidden generating component, you can ask for it
explicitly.

```{r}
sim_sampled_with_component <- simulate(
  model_sampled,
  params_df_sampled,
  keep_component = TRUE
)

table(sim_sampled_with_component$component)
```

For fixed mixtures, observed `R`/`rt` is enough. For sampled mixtures, the
current likelihood code still needs the per-trial component label, so we keep
that column here.

```{r}
data_sampled <- sim_sampled_with_component[, c("trial", "component", "R", "rt")]

prepared_sampled <- prepare_data(model_sampled, data_sampled)
ctx_sampled <- make_context(model_sampled)

params_df_sampled_true <- build_param_matrix(
  model_sampled,
  true_params_sampled,
  trial_df = prepared_sampled
)

ll_sampled_true <- as.numeric(log_likelihood(ctx_sampled, prepared_sampled, params_df_sampled_true))
ll_sampled_true
```

## Estimate a sampled-mixture weight
Latent weights can also be estimated. To keep the example simple, we estimate 
only the latent mixture weight `p_fast`, while holding the accumulator timing 
parameters fixed.

```{r eval=FALSE}
neg_loglik <- function(theta) {
  est <- true_params_sampled
  est["p_fast"] <- plogis(theta[["logit_p_fast"]])
  params_df <- build_param_matrix(
    model_sampled,
    est,
    trial_df = prepared_sampled
  )
  ll <- log_likelihood(ctx_sampled, prepared_sampled, params_df)
  -as.numeric(ll)
}

start <- c(logit_p_fast = qlogis(0.60))

set.seed(123456)
fit <- optim(start, neg_loglik, method = "BFGS")
```

```{r include=FALSE}
load_vignette_result <- function(name) {
  candidates <- c(
    file.path("vignettes", "data", paste0(name, ".RData")),
    file.path("data", paste0(name, ".RData"))
  )
  path <- candidates[file.exists(candidates)][1]
  load(path, envir = knitr::knit_global())
}

load_vignette_result("mixtures")
```

```{r}
fit_params <- c(p_fast = plogis(fit$par[["logit_p_fast"]]))
target <- c(p_fast = true_params_sampled[["p_fast"]])

data.frame(true = target, recovered = fit_params, miss = abs(target - fit_params))
```

Start with fixed components when trial type is known. The current
sampled-mixture likelihood implementation still expects the component column
during fitting, so this example estimates the sampled weight with that
component assignment retained.
