---
title: "Working with Mixtures"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with Mixtures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Here we show how mixtures work in `AccumulatR`.

Mixtures have two separate pieces:

- The model controls how mixture weights are defined. Use
  `set_mixture_options(mode = "fixed")` when the weights are known in
  advance, and `set_mixture_options(mode = "sample")` when one or more
  mixture weights should be estimated.
- The data control whether the component is observed on a trial. If a
  `component` column is present, likelihood evaluation conditions on that
  component. If `component` is omitted or set to `NA`, the likelihood and
  `response_probabilities()` marginalize over components.

```{r}
library(AccumulatR)
```

## A common architecture

Both sections use the same fast-versus-slow architecture. Response `R1` comes
from a target route and response `R2` comes from a competitor. The components
differ only in whether the target route is fast or slow on a given trial.

## Fixed mixture weights with observed components

In this first example, the mixture weights are fixed and the component label
is observed on each trial.

```{r}
model_fixed <- race_spec() |>
  add_accumulator("target_fast", "lognormal") |>
  add_accumulator("target_slow", "lognormal") |>
  add_accumulator("competitor", "lognormal") |>
  add_pool("TARGET", c("target_fast", "target_slow")) |>
  add_outcome("R1", "TARGET") |>
  add_outcome("R2", "competitor") |>
  add_component("fast", members = c("target_fast", "competitor"), weight = 0.4) |>
  add_component("slow", members = c("target_slow", "competitor"), weight = 0.6) |>
  set_mixture_options(mode = "fixed") |>
  finalize_model()

true_params_fixed <- c(
  target_fast.m = log(0.25),
  target_fast.s = 0.15,
  target_slow.m = log(0.45),
  target_slow.s = 0.20,
  competitor.m = log(0.35),
  competitor.s = 0.18
)
```

We now create a trial table that says which component each trial belongs to.

```{r}
set.seed(123456)

n_trials_fixed <- 1200
trial_types <- data.frame(
  trial = seq_len(n_trials_fixed),
  component = sample(c("fast", "slow"), n_trials_fixed, replace = TRUE, prob = c(0.4, 0.6)),
  stringsAsFactors = FALSE
)

params_df_fixed <- build_param_matrix(
  model_fixed,
  true_params_fixed,
  trial_df = trial_types
)

sim_fixed <- simulate(
  model_fixed,
  params_df_fixed,
  trial_df = trial_types,
  layout = "long"
)

data_fixed <- sim_fixed[, c("trial", "R", "rt", "component")]

table(data_fixed$component, data_fixed$R)
```

Because the component is known here, the simulated data keep that label. The
same pattern applies to real data: if trial type is observed, keep it in a
`component` column.

```{r}
prepared_fixed <- prepare_data(model_fixed, data_fixed)
ctx_fixed <- make_context(model_fixed)

params_df_fixed_true <- build_param_matrix(
  model_fixed,
  true_params_fixed,
  trial_df = prepared_fixed
)

ll_fixed <- as.numeric(log_likelihood(ctx_fixed, prepared_fixed, params_df_fixed_true))
ll_fixed
```

## Sampled mixture weights with latent components

Now we use the same architecture as a latent mixture. The component is not
observed per trial. Instead, the model samples a hidden component for each
trial, with `p_fast` controlling the proportion of fast trials.

```{r}
model_sampled <- race_spec() |>
  add_accumulator("target_fast", "lognormal") |>
  add_accumulator("target_slow", "lognormal") |>
  add_accumulator("competitor", "lognormal") |>
  add_pool("TARGET", c("target_fast", "target_slow")) |>
  add_outcome("R1", "TARGET") |>
  add_outcome("R2", "competitor") |>
  add_component("fast", members = c("target_fast", "competitor"), weight_param = "p_fast") |>
  add_component("slow", members = c("target_slow", "competitor")) |>
  set_mixture_options(mode = "sample", reference = "slow") |>
  finalize_model()

true_params_sampled <- c(
  target_fast.m = log(0.25),
  target_fast.s = 0.15,
  target_slow.m = log(0.45),
  target_slow.s = 0.20,
  competitor.m = log(0.35),
  competitor.s = 0.18,
  p_fast = 0.35
)
```

By default, sampled mixtures do not expose the component label in the output,
because that label is latent.

```{r}
set.seed(123456)

n_trials_sampled <- 1500
params_df_sampled <- build_param_matrix(
  model_sampled,
  true_params_sampled,
  n_trials = n_trials_sampled
)

sim_sampled <- simulate(model_sampled, params_df_sampled)
names(sim_sampled)
```

If you want to inspect the hidden generating component, you can ask for it
explicitly.

```{r}
sim_sampled_with_component <- simulate(
  model_sampled,
  params_df_sampled,
  keep_component = TRUE
)

table(sim_sampled_with_component$component)
```

For likelihood evaluation, we keep the sampled-mixture data in the same form
as real latent-mixture data: only the observed response `R` and response time
`rt`. Because `component` is omitted here, the likelihood marginalizes over
the possible components. If you do supply a `component` column, that
component is treated as observed on those trials.

```{r}
data_sampled <- sim_sampled[, c("trial", "R", "rt")]

prepared_sampled <- prepare_data(model_sampled, data_sampled)
ctx_sampled <- make_context(model_sampled)

params_df_sampled_true <- build_param_matrix(
  model_sampled,
  true_params_sampled,
  trial_df = prepared_sampled
)

ll_sampled_true <- as.numeric(log_likelihood(ctx_sampled, prepared_sampled, params_df_sampled_true))
ll_sampled_true
```

## Estimate a sampled-mixture weight
Latent weights can also be estimated. To keep the example simple, we estimate 
only the latent mixture weight `p_fast`, while holding the accumulator timing 
parameters fixed.

```{r eval=FALSE}
neg_loglik <- function(theta) {
  est <- true_params_sampled
  est["p_fast"] <- plogis(theta[["logit_p_fast"]])
  params_df <- build_param_matrix(
    model_sampled,
    est,
    trial_df = prepared_sampled
  )
  ll <- log_likelihood(ctx_sampled, prepared_sampled, params_df)
  -as.numeric(ll)
}

start <- c(logit_p_fast = qlogis(0.60))

set.seed(123456)
fit <- optim(start, neg_loglik, method = "BFGS")
```

```{r include=FALSE}
load_vignette_result <- function(name) {
  candidates <- c(
    file.path("vignettes", "data", paste0(name, ".RData")),
    file.path("data", paste0(name, ".RData"))
  )
  path <- candidates[file.exists(candidates)][1]
  load(path, envir = knitr::knit_global())
}

load_vignette_result("mixtures")
```

```{r}
fit_params <- c(p_fast = plogis(fit$par[["logit_p_fast"]]))
target <- c(p_fast = true_params_sampled[["p_fast"]])

data.frame(true = target, recovered = fit_params, miss = abs(target - fit_params))
```

Use `mode = "fixed"` when mixture weights are known and `mode = "sample"`
when mixture weights should be estimated. Independently of that choice,
include a `component` column only when the component is observed on a trial.
Otherwise omit it, or set it to `NA`, so the likelihood marginalizes over
components.
