---
title: "Simulate and Fit a Multi-Outcome Race Model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulate and Fit a Multi-Outcome Race Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette shows how to work with data in which more than one ordered
response is observed on each trial. Here each trial records the first response
(`R`, `rt`) and the second response (`R2`, `rt2`).

```{r}
library(AccumulatR)
```

**Define the model**
We use a simple two-accumulator race with direct responses `A` and `B`.
Setting `n_outcomes = 2` tells the model to retain the first and second
finishing responses.

```{r}
model_spec <- race_spec(n_outcomes = 2L) |>
  add_accumulator("A", "lognormal") |>
  add_accumulator("B", "lognormal") |>
  add_outcome("A", "A") |>
  add_outcome("B", "B")

structure <- finalize_model(model_spec)

true_params <- c(
  A.m = log(0.30),
  A.s = 0.18,
  A.q = 0,
  A.t0 = 0,
  B.m = log(0.38),
  B.s = 0.22,
  B.q = 0,
  B.t0 = 0
)
```

**Simulate data**
We generate data and keep both ordered responses for each trial.

```{r}
set.seed(123456)

n_trials <- 500
params_df <- build_param_matrix(model_spec, true_params, n_trials = n_trials)

sim <- simulate(structure, params_df)

data_df <- data.frame(
  trial = sim$trial,
  R = factor(sim$R),
  rt = sim$rt,
  R2 = factor(sim$R2),
  rt2 = sim$rt2,
  stringsAsFactors = FALSE
)

table(data_df$R, data_df$R2)
summary(data_df$rt2 - data_df$rt)
```

**Evaluate the likelihood**
We prepare the data once, build a model context, and evaluate the ranked
log-likelihood at the true parameter values.

```{r}
prepared <- prepare_data(structure, data_df)
ctx <- make_context(structure)

params_df_true <- build_param_matrix(
  model_spec,
  true_params,
  trial_df = prepared
)

ll_true <- as.numeric(log_likelihood(ctx, prepared, params_df_true))
ll_true
```

For comparison, we can also evaluate a parameter set that makes response `B`
too slow.

```{r}
wrong_params <- true_params
wrong_params["B.m"] <- log(0.55)

params_df_wrong <- build_param_matrix(
  model_spec,
  wrong_params,
  trial_df = prepared
)

ll_wrong <- as.numeric(log_likelihood(ctx, prepared, params_df_wrong))
ll_wrong
```

**Estimate parameters with `optim()`**
We estimate `A.m`, `A.s`, `B.m`, and `B.s`. The spread parameters are
optimized on the log scale and transformed back inside the objective
function.

```{r eval=FALSE}
neg_loglik <- function(theta) {
  est <- true_params
  est[c("A.m", "A.s", "B.m", "B.s")] <- theta[c("A.m", "A.s", "B.m", "B.s")]
  est[c("A.s", "B.s")] <- exp(est[c("A.s", "B.s")])
  params_df <- build_param_matrix(
    model_spec,
    est,
    trial_df = prepared
  )
  ll <- log_likelihood(ctx, prepared, params_df)
  -as.numeric(ll)
}

start <- c(
  A.m = log(0.24),
  A.s = log(0.12),
  B.m = log(0.48),
  B.s = log(0.12)
)

set.seed(123456)
fit <- optim(start, neg_loglik, method = "Nelder-Mead")
```

```{r include=FALSE}
load_vignette_result <- function(name) {
  candidates <- c(
    file.path("vignettes", "data", paste0(name, ".RData")),
    file.path("data", paste0(name, ".RData"))
  )
  path <- candidates[file.exists(candidates)][1]
  load(path, envir = knitr::knit_global())
}

load_vignette_result("multi_outcome")
```

```{r}
fit_params <- fit$par
fit_params[c("A.s", "B.s")] <- exp(fit_params[c("A.s", "B.s")])
target <- true_params[c("A.m", "A.s", "B.m", "B.s")]

data.frame(
  true = target,
  recovered = fit_params,
  miss = abs(target - fit_params)
)
```

The workflow is the same as in the single-response case, but the likelihood
now uses the second ranked response as additional information.
