---
title: "Simulate and Fit a Simple Race Model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulate and Fit a Simple Race Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette walks through the basic workflow for a race model: define the
architecture, simulate behavioral data, evaluate the likelihood, and recover
parameters. The example has two accumulators feeding response `R1` and one
accumulator feeding response `R2`.

```{r}
library(AccumulatR)
```

**Define the model**
We use three accumulators. `R1_A` and `R1_B` both support response `R1`, while
`R2` supports response `R2`.

```{r}
model_spec <- race_spec() |>
  add_accumulator("R1_A", "LBA") |>
  add_accumulator("R1_B", "RDM") |>
  add_accumulator("R2", "lognormal") |>
  add_pool("R1", c("R1_A", "R1_B")) |>
  add_outcome("R1", "R1") |>
  add_outcome("R2", "R2")

structure <- finalize_model(model_spec)

true_params <- c(
  R1_A.v = 2,
  R1_A.B = 1,
  R1_A.A = .3,
  R1_A.sv = 1,
  R1_A.q = 0,
  R1_A.t0 = 0,
  R1_B.v = 3,
  R1_B.B = 1,
  R1_B.A = .3,
  R1_B.s = 1,
  R1_B.q = 0,
  R1_B.t0 = 0,
  R2.m = log(0.4),
  R2.s = 0.18,
  R2.q = 0,
  R2.t0 = 0
)
```

**Simulate data**
We generate behavioral data consisting of responses and response times.

```{r}
set.seed(123456)

n_trials <- 2000
params_df <- build_param_matrix(model_spec, true_params, n_trials = n_trials)

sim <- simulate(structure, params_df)

data_df <- data.frame(
  trial = sim$trial,
  R = factor(sim$R),
  rt = sim$rt,
  stringsAsFactors = FALSE
)

table(data_df$R)
```

**Evaluate the likelihood**
We build a likelihood context and evaluate the log-likelihood of the simulated
behavioral data under the true parameter values.

```{r}
ctx <- build_likelihood_context(structure, data_df)

params_df_true <- build_param_matrix(
  model_spec,
  true_params,
  n_trials = max(data_df$trial),
  layout = ctx$param_layout
)

ll_true <- as.numeric(log_likelihood(ctx, params_df_true))
ll_true
```

We can compare that with a clearly misspecified parameter set.

```{r}
wrong_params <- true_params
wrong_params["R2.m"] <- log(0.45)

params_df_wrong <- build_param_matrix(
  model_spec,
  wrong_params,
  n_trials = max(data_df$trial),
  layout = ctx$param_layout
)

ll_wrong <- as.numeric(log_likelihood(ctx, params_df_wrong))
ll_wrong
```

**Estimate parameters with `optim()`**
We estimate five parameters: `R1_A.v`, `R1_B.v`, a shared threshold `B` used for both `R1_A.B` and `R1_B.B`, plus `R2.m` and `R2.s`. We estimate `B` and `s` on the log scale, to ensure support on the real line.

```{r}
neg_loglik <- function(theta) {
  est <- true_params
  est["R1_A.v"] <- theta[["R1_A.v"]]
  est["R1_B.v"] <- theta[["R1_B.v"]]
  est["R1_A.B"] <- exp(theta[["log_B_shared"]])
  est["R1_B.B"] <- exp(theta[["log_B_shared"]])
  est["R2.m"] <- theta[["R2.m"]]
  est["R2.s"] <- exp(theta[["log_R2.s"]])
  params_df <- build_param_matrix(
    model_spec,
    est,
    n_trials = max(data_df$trial),
    layout = ctx$param_layout
  )
  ll <- log_likelihood(ctx, params_df)
  -as.numeric(ll)
}

start <- c(
  R1_A.v = 1.5,
  R1_B.v = 1.5,
  log_B_shared = log(1.0),
  R2.m = log(0.32),
  log_R2.s = log(0.15)
)

fit <- optim(start, neg_loglik, method = "Nelder-Mead", control = list(maxit = 4000, reltol = 1e-9))
fit_params <- c(
  R1_A.v = fit$par[["R1_A.v"]],
  R1_B.v = fit$par[["R1_B.v"]],
  B_shared = exp(fit$par[["log_B_shared"]]),
  R2.m = fit$par[["R2.m"]],
  R2.s = exp(fit$par[["log_R2.s"]])
)
target <- c(
  R1_A.v = true_params[["R1_A.v"]],
  R1_B.v = true_params[["R1_B.v"]],
  B_shared = true_params[["R1_A.B"]],
  R2.m = true_params[["R2.m"]],
  R2.s = true_params[["R2.s"]]
)
data.frame(true = target, recovered = fit_params, miss = abs(target - fit_params))

```

That is the basic workflow the package is designed for: describe a response-time
model in psychological terms, simulate data, and then fit the same model back
to behavioral observations.
