---
title: "Trigger Types"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Trigger Types}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette shows how to add triggers to a model. Here `go1` and `go2`
produce responses `R1` and `R2`, but both accumulators are controlled by the
same trigger. On some trials that trigger fails, so no response is observed.

With `draw = "shared"`, the trigger failure is joint: both accumulators either
start together or fail together. With `draw = "independent"`, the same failure
probability is applied separately to each accumulator.

```{r}
library(AccumulatR)
```

**Define the model**
We use two lognormal accumulators and one shared trigger with failure
probability `q = 0.15`.

```{r}
model <- race_spec() |>
  add_accumulator("go1", "lognormal") |>
  add_accumulator("go2", "lognormal") |>
  add_outcome("R1", "go1") |>
  add_outcome("R2", "go2") |>
  add_trigger("shared_trigger",
    members = c("go1", "go2"),
    q = 0.15,
    draw = "shared"
  ) |>
  finalize_model()

true_params <- c(
  go1.m = log(0.30),
  go1.s = 0.18,
  go2.m = log(0.35),
  go2.s = 0.18
)
```

The trigger probability is fixed in the model definition here, so the fitted
parameter vector only contains the accumulator timing parameters.

**Simulate data**
Failed trigger trials appear as missing responses and missing response times.

```{r}
set.seed(123456)

n_trials <- 1500
params_df <- build_param_matrix(model, true_params, n_trials = n_trials)

sim <- simulate(model, params_df)

data_df <- data.frame(
  trial = sim$trial,
  R = factor(sim$R),
  rt = sim$rt,
  stringsAsFactors = FALSE
)

table(data_df$R, useNA = "ifany")
```

**Evaluate the likelihood**
We prepare the data, build a model context, and evaluate the shared-trigger
model under the true parameter values.

```{r}
prepared <- prepare_data(model, data_df)
ctx <- make_context(model)

params_df_true <- build_param_matrix(
  model,
  true_params,
  trial_df = prepared
)

ll_true <- as.numeric(log_likelihood(ctx, prepared, params_df_true))
ll_true
```

To see why the trigger type matters, compare that with the same model under
independent trigger failures instead of joint failures.

```{r}
model_independent <- race_spec() |>
  add_accumulator("go1", "lognormal") |>
  add_accumulator("go2", "lognormal") |>
  add_outcome("R1", "go1") |>
  add_outcome("R2", "go2") |>
  add_trigger("shared_trigger",
    members = c("go1", "go2"),
    q = 0.15,
    draw = "independent"
  ) |>
  finalize_model()

independent_prepared <- prepare_data(model_independent, data_df)
independent_ctx <- make_context(model_independent)

params_df_independent <- build_param_matrix(
  model_independent,
  true_params,
  trial_df = independent_prepared
)

ll_independent <- as.numeric(log_likelihood(
  independent_ctx,
  independent_prepared,
  params_df_independent
))
ll_independent
```

The same syntax can also be used when only one accumulator has a trigger.

```{r}
model_single_q <- race_spec() |>
  add_accumulator("go1", "lognormal") |>
  add_accumulator("go2", "lognormal") |>
  add_outcome("R1", "go1") |>
  add_outcome("R2", "go2") |>
  add_trigger("go1_trigger",
    members = c("go1"),
    q = 0.15
  ) |>
  finalize_model()
```

**Estimate parameters with `optim()`**
We estimate `go1.m`, `go1.s`, `go2.m`, and `go2.s`, while keeping the trigger
probability fixed at its generating value.

```{r eval=FALSE}
neg_loglik <- function(theta) {
  est <- true_params
  est["go1.m"] <- theta[["go1.m"]]
  est["go1.s"] <- exp(theta[["log_go1.s"]])
  est["go2.m"] <- theta[["go2.m"]]
  est["go2.s"] <- exp(theta[["log_go2.s"]])
  params_df <- build_param_matrix(
    model,
    est,
    trial_df = prepared
  )
  ll <- log_likelihood(ctx, prepared, params_df)
  -as.numeric(ll)
}

start <- c(
  go1.m = log(0.25),
  log_go1.s = log(0.12),
  go2.m = log(0.42),
  log_go2.s = log(0.12)
)

set.seed(123456)
fit <- optim(
  start,
  neg_loglik,
  method = "Nelder-Mead",
  control = list(maxit = 4000, reltol = 1e-9)
)
```

```{r include=FALSE}
load_vignette_result <- function(name) {
  candidates <- c(
    file.path("vignettes", "data", paste0(name, ".RData")),
    file.path("data", paste0(name, ".RData"))
  )
  path <- candidates[file.exists(candidates)][1]
  load(path, envir = knitr::knit_global())
}

load_vignette_result("trigger_model")
```

```{r}
fit_params <- c(
  go1.m = fit$par[["go1.m"]],
  go1.s = exp(fit$par[["log_go1.s"]]),
  go2.m = fit$par[["go2.m"]],
  go2.s = exp(fit$par[["log_go2.s"]])
)
target <- true_params[c("go1.m", "go1.s", "go2.m", "go2.s")]

data.frame(true = target, recovered = fit_params, miss = abs(target - fit_params))
```

Use shared triggers when several accumulators are governed by the same gating
event. Use independent triggers when the same failure probability should apply
separately to each accumulator.
