---
title: "Stim-Selective Stop Likelihood"
output:
  html_document: default
  pdf_document: default
---

This note derives a hand-written likelihood for the stop component of the stimulus-selective stop model and compares it directly to `AccumulatR`.

```{r}
library(AccumulatR)
model <- race_spec() |>
  add_accumulator("A", "lognormal") |>
  add_accumulator("B", "lognormal") |>
  add_accumulator("S1", "lognormal") |>
  add_accumulator("IS", "lognormal") |>
  add_accumulator("S2", "lognormal") |>
  add_outcome(
    "A",
    first_of(
      inhibit("A", by = "S1"),
      all_of("A", "S1", inhibit("IS", by = "S2"))
    )
  ) |>
  add_outcome(
    "B",
    first_of(
      inhibit("B", by = "S1"),
      all_of("B", "S1", inhibit("IS", by = "S2"))
    )
  ) |>
  add_outcome(
    "STOP",
    all_of("S1", inhibit("S2", by = "IS")),
    options = list(map_outcome_to = NA_character_)
  ) |>
  add_component("go", members = c("A", "B")) |>
  add_component("stop", members = c("A", "B", "S1", "IS", "S2")) |>
  set_parameters(list(
    m_go = c("A.m", "B.m"),
    s_go = c("A.s", "B.s"),
    t0_go = c("A.t0", "B.t0")
  )) |>
  finalize_model()
```

Under the `stop` component there are three declared outcomes:

-   `A = first_of(inhibit(A, S1), all_of(A, S1, inhibit(IS, S2)))`
-   `B = first_of(inhibit(B, S1), all_of(B, S1, inhibit(IS, S2)))`
-   `STOP = all_of(S1, inhibit(S2, IS))`, mapped to `NA`

Consider the two branches of A and B:

-   early `A`: `E_A = inhibit(A, S1)`
-   late `A`: `L_A = all_of(A, S1, inhibit(IS, S2))`
-   early `B`: `E_B = inhibit(B, S1)`
-   late `B`: `L_B = all_of(B, S1, inhibit(IS, S2))`

Then:

-   `A_out = first_of(E_A, L_A)`
-   `B_out = first_of(E_B, L_B)`

For the current engine, observed `NA` is leftover visible mass after `A` and `B`, so the hand-written likelihood targets:

-   `p(A_obs at t) = f_A_out(t) S_B_out(t)`
-   `p(B_obs at t) = f_B_out(t) S_A_out(t)`
-   `P(NA_obs) = 1 - P(A_obs) - P(B_obs)`

## The actual formulas

Write `f_X(t)`, `F_X(t)`, and `S_X(t)` for the density, cdf, and survival of accumulator `X`.

For this model:

-   `E_A = inhibit(A, S1)`
-   `E_B = inhibit(B, S1)`
-   `H = inhibit(IS, S2)`
-   `L_A = all_of(A, S1, H)`
-   `L_B = all_of(B, S1, H)`

The pieces are:

-   `f_EA(t) = f_A(t) S_S1(t)`

-   `f_EB(t) = f_B(t) S_S1(t)`

-   `f_H(t) = f_IS(t) S_S2(t)`

-   `F_EA(t) = ∫_0^t f_A(u) S_S1(u) du`

-   `F_EB(t) = ∫_0^t f_B(u) S_S1(u) du`

-   `F_H(t) = ∫_0^t f_IS(u) S_S2(u) du`

-   `S_EA(t) = 1 - F_EA(t)`

-   `S_EB(t) = 1 - F_EB(t)`

-   `S_H(t) = 1 - F_H(t)`

For the late branches:

-   `F_LA(t) = F_A(t) F_S1(t) F_H(t)`

-   `F_LB(t) = F_B(t) F_S1(t) F_H(t)`

-   `f_LA(t) = f_A(t) F_S1(t) F_H(t) + F_A(t) f_S1(t) F_H(t) + F_A(t) F_S1(t) f_H(t)`

-   `f_LB(t) = f_B(t) F_S1(t) F_H(t) + F_B(t) f_S1(t) F_H(t) + F_B(t) F_S1(t) f_H(t)`

-   `S_LA(t) = 1 - F_LA(t)`

-   `S_LB(t) = 1 - F_LB(t)`

The visible outcomes are:

-   `A_out = first_of(E_A, L_A)`
-   `B_out = first_of(E_B, L_B)`

So:

-   `f_A_out(t) = f_EA(t) S_LA(t) + f_LA(t) S_EA(t)`

-   `f_B_out(t) = f_EB(t) S_LB(t) + f_LB(t) S_EB(t)`

-   `S_A_out(t) = S_EA(t) S_LA(t)`

-   `S_B_out(t) = S_EB(t) S_LB(t)`

That gives the observed response densities directly:

-   `p(A_obs at t) = f_A_out(t) S_B_out(t)`
-   `p(B_obs at t) = f_B_out(t) S_A_out(t)`

Expanding `A` into its two branches:

-   early `A` term:

    `f_EA(t) S_LA(t) S_EB(t) S_LB(t)`

-   late `A` term:

    `f_LA(t) S_EA(t) S_EB(t) S_LB(t)`

and the total is their sum.

By symmetry:

-   early `B` term:

    `f_EB(t) S_LB(t) S_EA(t) S_LA(t)`

-   late `B` term:

    `f_LB(t) S_EB(t) S_EA(t) S_LA(t)`

and again the total is their sum.

This is the R code

```{r}
params <- c(
  m_go = log(0.30), s_go = 0.18, t0_go = 0.05,
  S1.m = log(0.26), S1.s = 0.18, S1.t0 = 0.00,
  IS.m = log(0.35), IS.s = 0.18, IS.t0 = 0.00,
  S2.m = log(0.32), S2.s = 0.18, S2.t0 = 0.00
)

go_parts <- function(params) {
  list(m = params[["m_go"]], s = params[["s_go"]], q = 0, t0 = params[["t0_go"]])
}

acc_parts <- function(prefix, params) {
  list(
    m = params[[paste0(prefix, ".m")]],
    s = params[[paste0(prefix, ".s")]],
    q = 0,
    t0 = params[[paste0(prefix, ".t0")]]
  )
}

acc_pdf <- function(t, p) {
  if (!is.finite(t) || t < p$t0) return(0)
  (1 - p$q) * dlnorm(t - p$t0, meanlog = p$m, sdlog = p$s)
}

acc_cdf <- function(t, p) {
  if (is.infinite(t) && t > 0) return(1 - p$q)
  if (!is.finite(t) || t < p$t0) return(0)
  (1 - p$q) * plnorm(t - p$t0, meanlog = p$m, sdlog = p$s)
}

acc_surv <- function(t, p) 1 - acc_cdf(t, p)

# Return density/cdf/survival for one primitive accumulator.
leaf <- function(p) {
  force(p)
  function(t) {
    list(
      density = acc_pdf(t, p),
      cdf = acc_cdf(t, p),
      survival = acc_surv(t, p)
    )
  }
}

# guard(R, B): R succeeds while B has not yet finished.
guard <- function(reference, blocker) {
  function(t) {
    if (!is.finite(t) || t < 0) {
      return(list(density = 0, cdf = 0, survival = 1))
    }
    density <- reference(t)$density * blocker(t)$survival
    cdf <- if (t <= 0) 0 else {
      integrate(
        function(u) {
          vapply(u, function(uu) reference(uu)$density * blocker(uu)$survival, numeric(1))
        },
        lower = 0,
        upper = t,
        rel.tol = 1e-8,
        abs.tol = 1e-10,
        subdivisions = 200L
      )$value
    }
    cdf <- min(max(cdf, 0), 1)
    list(density = density, cdf = cdf, survival = 1 - cdf)
  }
}

# all_of(X1, X2, ..., Xn)
all_of_node <- function(...) {
  children <- list(...)
  function(t) {
    vals <- lapply(children, function(fn) fn(t))
    Fvals <- vapply(vals, `[[`, numeric(1), "cdf")
    fvals <- vapply(vals, `[[`, numeric(1), "density")
    cdf <- prod(Fvals)
    density <- sum(vapply(
      seq_along(fvals),
      function(i) fvals[i] * if (length(fvals) == 1L) 1 else prod(Fvals[-i]),
      numeric(1)
    ))
    cdf <- min(max(cdf, 0), 1)
    list(density = density, cdf = cdf, survival = 1 - cdf)
  }
}

# first_of(X1, X2, ..., Xn)
first_of_node <- function(...) {
  children <- list(...)
  function(t) {
    vals <- lapply(children, function(fn) fn(t))
    Svals <- vapply(vals, `[[`, numeric(1), "survival")
    fvals <- vapply(vals, `[[`, numeric(1), "density")
    survival <- prod(Svals)
    density <- sum(vapply(
      seq_along(fvals),
      function(i) fvals[i] * if (length(fvals) == 1L) 1 else prod(Svals[-i]),
      numeric(1)
    ))
    survival <- min(max(survival, 0), 1)
    list(density = density, cdf = 1 - survival, survival = survival)
  }
}

A  <- leaf(go_parts(params))
B  <- leaf(go_parts(params))
S1 <- leaf(acc_parts("S1", params))
IS <- leaf(acc_parts("IS", params))
S2 <- leaf(acc_parts("S2", params))

early_A <- guard(A, S1)
early_B <- guard(B, S1)
ignore_gate <- guard(IS, S2)
late_A <- all_of_node(A, S1, ignore_gate)
late_B <- all_of_node(B, S1, ignore_gate)

A_out <- first_of_node(early_A, late_A)
B_out <- first_of_node(early_B, late_B)

# Exact observed response density under the stop component.
obs_density <- function(response, t) {
  if (response == "A") {
    A_out(t)$density * B_out(t)$survival
  } else {
    B_out(t)$density * A_out(t)$survival
  }
}

# Exact observed response probabilities.
obs_prob <- function(response) {
  integrate(
    function(t) vapply(t, function(tt) obs_density(response, tt), numeric(1)),
    lower = 0,
    upper = Inf,
    rel.tol = 1e-7,
    abs.tol = 1e-9,
    subdivisions = 200L,
    stop.on.error = FALSE
  )$value
}

# Engine-matching NA probability.
na_prob <- function() {
  1 - obs_prob("A") - obs_prob("B")
}
```

## Agreement with the engine

The comparison below uses the same parameter vector for both the hand-written likelihood and the engine.

```{r}
ctx <- make_context(model)

engine_likelihood <- function(response, rt) {
  data_df <- data.frame(
    trial = 1L,
    component = "stop",
    R = if (is.na(response)) NA_character_ else response,
    rt = rt,
    stringsAsFactors = FALSE
  )
  prepared <- prepare_data(model, data_df)
  params_df <- build_param_matrix(model$model_spec, params, trial_df = prepared)
  exp(as.numeric(log_likelihood(ctx, prepared, params_df, min_ll = -1e12)))
}

comparison <- data.frame(
  component = "stop",
  response = c("A", "B", NA_character_),
  rt = c(0.42, 0.47, NA_real_),
  hand_written = c(
    obs_density("A", 0.42),
    obs_density("B", 0.47),
    na_prob()
  ),
  accumulatr = c(
    engine_likelihood("A", 0.42),
    engine_likelihood("B", 0.47),
    engine_likelihood(NA_character_, NA_real_)
  ),
  stringsAsFactors = FALSE
)

comparison$diff <- comparison$accumulatr - comparison$hand_written
comparison
```

For this model, the engine-matching target is exactly:

-   `p(A_obs at t) = f_A_out(t) S_B_out(t)`
-   `p(B_obs at t) = f_B_out(t) S_A_out(t)`
-   `P(NA_obs) = 1 - P(A_obs) - P(B_obs)`
