---
title: "Stop-Change Likelihood"
output:
  html_document: default
  pdf_document: default
---

This note derives a hand-written likelihood for the stop-change model and compares it directly to `AccumulatR`.

```{r}
library(AccumulatR)

model <- race_spec() |>
  add_accumulator("S", "lognormal") |>
  add_accumulator("stop", "lognormal") |>
  add_accumulator("change", "lognormal") |>
  add_outcome("S", inhibit("S", by = "stop")) |>
  add_outcome("X", all_of("change", "stop")) |>
  add_component("go_only", members = "S", weight = .75) |>
  add_component("go_stop", members = c("S", "stop", "change"), weight = .25) |>
  add_trigger(
    "stop_trigger",
    members = c("stop", "change"),
    q = 0.05,
    param = "stop_trigger"
  ) |>
  set_mixture_options(mode = "fixed") |>
  set_parameters(list(
    m_go = "S.m",
    m_stop = "stop.m",
    m_change = "change.m",
    s_go = "S.s",
    s_stop = "stop.s",
    s_change = "change.s",
    t0_go = "S.t0",
    t0_change = "change.t0",
    q = c("stop.q", "change.q")
  )) |>
  finalize_model()
```

Under the `go_stop` component there are two declared outcomes:

-   `S = inhibit(S, stop)`
-   `X = all_of(change, stop)`

The trigger is shared across `stop` and `change`. Inside `go_stop` there are two trigger states:

-   fail, with probability `q`: both `stop` and `change` are absent
-   succeed, with probability `1 - q`: both are active

For the intended model, the hand-written component-conditioned targets are:

-   `p(S at t | go_only) = f_S(t)`
-   `p(S at t | go_stop) = q f_S(t) + (1 - q) f_S(t) S_stop(t)`
-   `p(X at t | go_stop) = (1 - q) f_X_out(t) S_S_out(t)`

where:

-   `f_X_out(t) = f_change(t) F_stop(t) + F_change(t) f_stop(t)`
-   `F_S_out(t) = ∫_0^t f_S(u) S_stop(u) du`
-   `S_S_out(t) = 1 - F_S_out(t)`

## The actual formulas

Write `f_S`, `F_S`, and `S_S` for the density, cdf, and survival of accumulator `S`, and similarly for `stop` and `change`.

For this model:

-   `S_out = inhibit(S, stop)`
-   `X_out = all_of(change, stop)`

So:

-   `f_S_out(t) = f_S(t) S_stop(t)`
-   `F_S_out(t) = ∫_0^t f_S(u) S_stop(u) du`
-   `S_S_out(t) = 1 - F_S_out(t)`
-   `F_X_out(t) = F_change(t) F_stop(t)`
-   `f_X_out(t) = f_change(t) F_stop(t) + F_change(t) f_stop(t)`

That gives the intended component-conditioned densities:

-   `p(S at t | go_only) = f_S(t)`
-   `p(S at t | go_stop) = q f_S(t) + (1 - q) f_S(t) S_stop(t)`
-   `p(X at t | go_stop) = (1 - q) f_X_out(t) S_S_out(t)`

This is the R code.

```{r}
params <- c(
  m_go = log(0.30),
  s_go = 0.18,
  t0_go = 0.00,
  m_stop = log(0.22),
  s_stop = 0.18,
  m_change = log(0.40),
  s_change = 0.18,
  t0_change = 0.00,
  q = 0.05,
  S.q = 0.00
)

ctx <- make_context(model)

acc_pdf <- function(t, m, s, t0 = 0, q = 0) {
  out <- numeric(length(t))
  ok <- is.finite(t) & t >= t0
  out[ok] <- (1 - q) * dlnorm(t[ok] - t0, meanlog = m, sdlog = s)
  out
}

acc_cdf <- function(t, m, s, t0 = 0, q = 0) {
  out <- numeric(length(t))
  out[is.infinite(t) & t > 0] <- 1 - q
  ok <- is.finite(t) & t >= t0
  out[ok] <- (1 - q) * plnorm(t[ok] - t0, meanlog = m, sdlog = s)
  pmin(pmax(out, 0), 1)
}

f_S <- function(t) {
  acc_pdf(t, params[["m_go"]], params[["s_go"]], params[["t0_go"]], q = 0)
}

F_stop <- function(t) {
  acc_cdf(t, params[["m_stop"]], params[["s_stop"]], 0, q = 0)
}

f_stop <- function(t) {
  acc_pdf(t, params[["m_stop"]], params[["s_stop"]], 0, q = 0)
}

F_change <- function(t) {
  acc_cdf(t, params[["m_change"]], params[["s_change"]], params[["t0_change"]], q = 0)
}

f_change <- function(t) {
  acc_pdf(t, params[["m_change"]], params[["s_change"]], params[["t0_change"]], q = 0)
}

F_S_out_1 <- function(t) {
  if (!is.finite(t) || t <= 0) {
    return(0)
  }
  integrate(
    function(u) f_S(u) * (1 - F_stop(u)),
    lower = 0,
    upper = t,
    rel.tol = 1e-10,
    abs.tol = 1e-12,
    subdivisions = 500L
  )$value
}

S_S_out <- function(t) {
  1 - vapply(t, F_S_out_1, numeric(1))
}

q_trig <- 0.05

hand_go_only_S <- function(t) {
  f_S(t)
}

hand_go_stop_S <- function(t) {
  q_trig * f_S(t) + (1 - q_trig) * f_S(t) * (1 - F_stop(t))
}

hand_go_stop_X <- function(t) {
  f_X_out <- f_change(t) * F_stop(t) + F_change(t) * f_stop(t)
  (1 - q_trig) * f_X_out * S_S_out(t)
}

engine_trial <- function(component, R, rt) {
  data_df <- data.frame(
    trial = 1L,
    component = component,
    R = R,
    rt = rt,
    stringsAsFactors = FALSE
  )
  prepared <- prepare_data(model, data_df)
  params_df <- build_param_matrix(model$model_spec, params, trial_df = prepared)
  exp(log_likelihood(ctx, prepared, params_df))
}

comparison <- data.frame(
  case = c("go_only, S @ 0.30", "go_stop, S @ 0.30", "go_stop, X @ 0.50"),
  hand_written = c(
    hand_go_only_S(0.30),
    hand_go_stop_S(0.30),
    hand_go_stop_X(0.50)
  ),
  accumulatr = c(
    engine_trial("go_only", "S", 0.30),
    engine_trial("go_stop", "S", 0.30),
    engine_trial("go_stop", "X", 0.50)
  )
)

comparison$diff <- comparison$hand_written - comparison$accumulatr
comparison
```

`go_only -> S` matches the engine exactly. `go_stop -> X` also matches. `go_stop -> S` does not.

The mismatch is not coming from the hand derivation being numerically unstable. It is structural. Numerically, the current engine is closely tracked by

-   `f_S(t) S_change(t) [q + (1 - q) S_stop(t)]`

rather than by the intended outcome-level expression

-   `q f_S(t) + (1 - q) f_S(t) S_stop(t)`

So the current engine appears to be carrying `change` as an extra survival constraint on `S`.
